Appendix 2: Simulating 20-species landscapes

### 2016-02-22 Harris_2015/manuscript-materials/draft.Rmd
cleanup

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, based on controlling for indirect effects with partial
correlations or generalized linear models, also performed well. Null models 
showed no evidence of being able to control for indirect effects, and reliably 
yielded incorrect inferences when such effects were present.

@@

Following @azaele_inferring_2010, this paper shows that Markov networks
[undirected graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding the
landscape-level consequences of pairwise species interactions,
and for estimating them from observed presence-absence matrices. Markov networks
have been used in diverse scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it determines the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates would require normalizing the relative 
probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized, one can find exact maximum likelihood estimates for
$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like 
the ones considered here have unimodal likelihood surfaces 
[@murphy_machine_2012], so these estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation 1 
can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
they want to fit a Markov network model. For the analyses presented here, I used 
the rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its gradients
exactly [@murphy_machine_2012]; the package passes these functions to the "BFGS" 
method in R's general-purpose optimizer, which finds values for $\alpha$ and 
$\beta$.

@@

The next set of landscapes provided a way to assess each method's ability to 
identify direct interactions in the presence of environmental heterogeneity.
Here, the $\beta$ coefficients were calculated as above, but each species'
$\alpha$ value depended linearly on two normally-distributed environmental 
factors, which were drawn from independent Gaussians for each site. Once the 
local $\alpha$ values were calculated, independent Gibbs samplers determined 
the species composition for each site based on Equation 1.

In the final set of landscapes, I simulated each species' abundance (instead of
just their presence/absence); here, interactions between species occurred on a 
per-capita basis. To prevent runaway mutualisms leading to infinite abundance,
all interaction coefficients were negative in these simulations. Good 
performance on these landscapes would indicate some robustness to the
mechanistic details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a 
weakly-informative regularizer (equivalent to a logistic prior with location 0
and scale 2) ensured that the estimates were always finite.

@@

The next method used the Pairs software described in @gotelli_empirical_2009.
This program simulating new landscapes from a null model that retains the row 
and column sums of the original matrix [@strong_ecological_1984] and calculates
$Z$-scores to summarize a species pair's deviation from this null.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: 0.02 for simulations that matched the model's assumptions,
versus 0.14 for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were 0.31 and 0.53 for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than 5 times as often as the Markov network, even
though it only rejects the null hypothesis twice as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, so it 
never misclassified strong interactions like the null model did (Figure 4C).

@@

\noindent
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from species composition data, even for cases where
environmental heterogeneity and indirect interactions cause ecologists' typical
null modeling approaches to reliably fail. Partial correlations and generalized
linear models can both provide computationally efficient approximations, but
with somewhat lower accuracy [especially for typically-sized data sets with small
numbers of sites; @gotelli_empirical_2009]. The difference in accuracy may be
larger for real data sets than for the simulated landscapes in Figure 3,
however; linear approximations to the Markov network make larger errors when
the interaction matrix is structured [e.g. due to guilds or trophic levels;
@loh_structure_2013]. Similarly, the separate generalized linear models
for each species can severely overfit in some cases [@lee_learning_2012]. The
full Markov network should thus be preferred to the approximations when
it is computationally tractable.

>>>>

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null models 
showed no evidence of being able to control for indirect effects, and reliably 
yielded incorrect inferences when such effects were present.

@@

Following @azaele_inferring_2010, this paper shows that Markov networks
[undirected graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding the
landscape-level consequences of pairwise species interactions,
and for estimating them from observed presence-absence matrices. Markov networks
have been used in diverse scientific fields for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it determines the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized (Figure 2B), one can find exact maximum likelihood
estimates for $\alpha$ and $\beta$ by numerically optimizing
$p(\vec{y}|\alpha, \beta)$. Fully-observed networks like the ones considered 
here have unimodal likelihood surfaces [@murphy_machine_2012], so these
estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation 1 
can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
they want to fit a Markov network model. For the analyses presented here, where
the number of species did not exceed 20, approximations were not necessary. Instead,
I used the rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its gradients
exactly [@murphy_machine_2012]; the package passes these functions to the "BFGS" 
method in R's general-purpose optimizer, which finds values for $\alpha$ and 
$\beta$.

@@

The next set of landscapes provided a way to assess each method's ability to 
identify direct interactions in the presence of environmental heterogeneity.
Here, the $\beta$ coefficients were calculated as above, but each species'
$\alpha$ value depended linearly on two environmental factors, which were drawn 
from independent Gaussians for each site. Once the local $\alpha$ values were
calculated, independent Gibbs samplers determined the species composition for 
each site based on Equation 1.

In the final set of landscapes, I simulated each species' abundance (instead of
just presence/absence); furthermore, interactions between species occurred on a 
per-capita basis in these simulations. To prevent runaway mutualisms leading to
infinite abundance, all interaction coefficients were negative in these 
simulations. Good performance on these landscapes would indicate some robustness 
to the mechanistic details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a 
weakly-informative regularizer (equivalent to a logistic prior with location 0
and scale 2) ensured that the estimates were always finite (Appendix 2).

@@

The next method used the Pairs software described in @gotelli_empirical_2009.
This program simulates new landscapes from a null model that retains the row 
and column sums of the original matrix [@strong_ecological_1984] and calculates
$Z$-scores to summarize a species pair's deviation from this null.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: 0.02 for simulations that matched the model's assumptions,
versus 0.14 for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were 0.31 and 0.53 for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than 5 times as often as the Markov network, even
though it only rejects the null hypothesis twice as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, and it 
never misclassified strong interactions like the null model did (Figure 4C).

@@

\noindent
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from species composition data, even for cases where
environmental heterogeneity and indirect interactions cause ecologists' typical
null modeling approaches to reliably fail. Partial correlations and generalized
linear models can both provide computationally efficient approximations, but
with somewhat lower accuracy [especially for typically-sized data sets with small
numbers of sites; @gotelli_empirical_2009]. The difference in accuracy may be
even larger for real data sets than for the simulated landscapes in Figure 3; 
linear approximations to the Markov network make larger errors when
the interaction matrix is structured [e.g. due to guilds or trophic levels;
@loh_structure_2013]. Similarly, the separate generalized linear models
for each species can severely overfit in some cases [@lee_learning_2012]. The
full Markov network should thus be preferred to the approximations when
it is computationally tractable.

### 2016-02-22 Harris_2015/manuscript-materials/draft.tex
cleanup

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. The Markov networks
consistently outperformed other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, based on controlling for indirect effects with partial
correlations or generalized linear models, also performed well. Null
models showed no evidence of being able to control for indirect effects,
and reliably yielded incorrect inferences when such effects were
present.

@@

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for estimating them
from observed presence-absence matrices. Markov networks have been used
in diverse scientific fields in similar contexts for decades, from
physics (where nearby particles interact magnetically; Cipra 1987) to
spatial statistics (where adjacent grid cells have correlated values;
Harris 1974, Gelfand et al. 2005). While community ecologists explored
some related approaches in the 1980's (Whittam and Siegel-Causey 1981),
they used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it determines the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = +2\), then each species' odds of
occurrence would be \(e^2\) times higher when the other one is present
(as compared with otherwise equivalent sites). The relative probability
of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species
co-occur. As a result, the model tends---all else equal---to produce
assemblages where many positively-associated species pairs co-occur and
few negatively-associated pairs do (just as an ecologist might expect).
When all else is \emph{not} equal (e.g.~Figure 1, where the presence of
one competitor is associated with release from another competitor), then
predicting species' overall co-occurrence rates would require
normalizing the relative probabilities of possible \(\vec{y}\) vectors
so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized, one can find exact maximum likelihood estimates for
\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), so these estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation
1 can become intractable, and researchers will either need to
approximate it (Lee and Hastie 2012) or approximate its gradient (Harris
2016) if they want to fit a Markov network model. For the analyses
presented here, I used the rosalia package (Harris 2015a) for the R
programming language (R Core Team 2015) to calculate
\(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
the package passes these functions to the ``BFGS'' method in R's
general-purpose optimizer, which finds values for \(\alpha\) and
\(\beta\).

@@

The next set of landscapes provided a way to assess each method's
ability to identify direct interactions in the presence of environmental
heterogeneity. Here, the \(\beta\) coefficients were calculated as
above, but each species' \(\alpha\) value depended linearly on two
normally-distributed environmental factors, which were drawn from
independent Gaussians for each site. Once the local \(\alpha\) values
were calculated, independent Gibbs samplers determined the species
composition for each site based on Equation 1.

In the final set of landscapes, I simulated each species' abundance
(instead of just their presence/absence); here, interactions between
species occurred on a per-capita basis. To prevent runaway mutualisms
leading to infinite abundance, all interaction coefficients were
negative in these simulations. Good performance on these landscapes
would indicate some robustness to the mechanistic details of species
interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a
weakly-informative regularizer (equivalent to a logistic prior with
location 0 and scale 2) ensured that the estimates were always finite.

@@

The next method used the Pairs software described in Gotelli and Ulrich
(2009). This program simulating new landscapes from a null model that
retains the row and column sums of the original matrix (Strong et al.
1984) and calculates \(Z\)-scores to summarize a species pair's
deviation from this null.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: 0.02 for simulations that matched the model's
assumptions, versus 0.14 for simulations that included environmental
heterogeneity (see Appendix 4 for confidence interval coverage across a
range of \(\beta_{ij}\) values). The null model's Type I error rates
were 0.31 and 0.53 for the constant and heterogeneous landscapes,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the
interaction with 95\% confidence. The null model makes such errors more
than 5 times as often as the Markov network, even though it only rejects
the null hypothesis twice as often overall (Appendix 4). The Markov
network's errors were also more concentrated around 0, so it never
misclassified strong interactions like the null model did (Figure 4C).

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from species composition data,
even for cases where environmental heterogeneity and indirect
interactions cause ecologists' typical null modeling approaches to
reliably fail. Partial correlations and generalized linear models can
both provide computationally efficient approximations, but with somewhat
lower accuracy (especially for typically-sized data sets with small
numbers of sites; Gotelli and Ulrich 2009). The difference in accuracy
may be larger for real data sets than for the simulated landscapes in
Figure 3, however; linear approximations to the Markov network make
larger errors when the interaction matrix is structured (e.g.~due to
guilds or trophic levels; Loh and Wainwright 2013). Similarly, the
separate generalized linear models for each species can severely overfit
in some cases (Lee and Hastie 2012). The full Markov network should thus
be preferred to the approximations when it is computationally tractable.

>>>>

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. The Markov networks
consistently outperformed other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null
models showed no evidence of being able to control for indirect effects,
and reliably yielded incorrect inferences when such effects were
present.

@@

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for estimating them
from observed presence-absence matrices. Markov networks have been used
in diverse scientific fields for decades, from physics (where nearby
particles interact magnetically; Cipra 1987) to spatial statistics
(where adjacent grid cells have correlated values; Harris 1974, Gelfand
et al. 2005). While community ecologists explored some related
approaches in the 1980's (Whittam and Siegel-Causey 1981), they used
severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it determines the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = +2\), then each species' odds of
occurrence would be \(e^2\) times higher when the other one is present
(as compared with otherwise equivalent sites). The relative probability
of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species
co-occur. As a result, the model tends---all else equal---to produce
assemblages where many positively-associated species pairs co-occur and
few negatively-associated pairs do (just as an ecologist might expect).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized (Figure 2B), one can find exact maximum likelihood estimates
for \(\alpha\) and \(\beta\) by numerically optimizing
\(p(\vec{y}|\alpha, \beta)\). Fully-observed networks like the ones
considered here have unimodal likelihood surfaces (Murphy 2012), so
these estimates will represent the global optimum.

When the number of species is larger than about 30, noralizing Equation
1 can become intractable, and researchers will either need to
approximate it (Lee and Hastie 2012) or approximate its gradient (Harris
2016) if they want to fit a Markov network model. For the analyses
presented here, where the number of species did not exceed 20,
approximations were not necessary. Instead, I used the rosalia package
(Harris 2015a) for the R programming language (R Core Team 2015) to
calculate \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly
(Murphy 2012); the package passes these functions to the ``BFGS'' method
in R's general-purpose optimizer, which finds values for \(\alpha\) and
\(\beta\).

@@

The next set of landscapes provided a way to assess each method's
ability to identify direct interactions in the presence of environmental
heterogeneity. Here, the \(\beta\) coefficients were calculated as
above, but each species' \(\alpha\) value depended linearly on two
environmental factors, which were drawn from independent Gaussians for
each site. Once the local \(\alpha\) values were calculated, independent
Gibbs samplers determined the species composition for each site based on
Equation 1.

In the final set of landscapes, I simulated each species' abundance
(instead of just presence/absence); furthermore, interactions between
species occurred on a per-capita basis in these simulations. To prevent
runaway mutualisms leading to infinite abundance, all interaction
coefficients were negative in these simulations. Good performance on
these landscapes would indicate some robustness to the mechanistic
details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a
weakly-informative regularizer (equivalent to a logistic prior with
location 0 and scale 2) ensured that the estimates were always finite
(Appendix 2).

@@

The next method used the Pairs software described in Gotelli and Ulrich
(2009). This program simulates new landscapes from a null model that
retains the row and column sums of the original matrix (Strong et al.
1984) and calculates \(Z\)-scores to summarize a species pair's
deviation from this null.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: 0.02 for simulations that matched the model's
assumptions, versus 0.14 for simulations that included environmental
heterogeneity (see Appendix 4 for confidence interval coverage across a
range of \(\beta_{ij}\) values). The null model's Type I error rates
were 0.31 and 0.53 for the constant and heterogeneous landscapes,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the
interaction with 95\% confidence. The null model makes such errors more
than 5 times as often as the Markov network, even though it only rejects
the null hypothesis twice as often overall (Appendix 4). The Markov
network's errors were also more concentrated around 0, and it never
misclassified strong interactions like the null model did (Figure 4C).

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from species composition data,
even for cases where environmental heterogeneity and indirect
interactions cause ecologists' typical null modeling approaches to
reliably fail. Partial correlations and generalized linear models can
both provide computationally efficient approximations, but with somewhat
lower accuracy (especially for typically-sized data sets with small
numbers of sites; Gotelli and Ulrich 2009). The difference in accuracy
may be even larger for real data sets than for the simulated landscapes
in Figure 3; linear approximations to the Markov network make larger
errors when the interaction matrix is structured (e.g.~due to guilds or
trophic levels; Loh and Wainwright 2013). Similarly, the separate
generalized linear models for each species can severely overfit in some
cases (Lee and Hastie 2012). The full Markov network should thus be
preferred to the approximations when it is computationally tractable.

### 2016-02-22 Harris_2015/simulation.Rmd
more manuscript and appendices
cleanup

The following function draws random coefficients for a Markov network of a pre-specified size.

The $alpha$ coefficients controlling species' intercepts were drawn from 
Gaussian distributions with unit variance and a mean given by `mean_alpha`.
In the R code at the end of this document, `mean_alpha` was defined as $-1$ for presence-absence simulations and $5$ for the abundance simulations.

@@

```{r, cache=TRUE}
make_coefficients = function(n_spp, p_neg, mean_alpha){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

>>>>

The following function draws random coefficients for a Markov network of a pre-specified size (20 species in these examples).

The $alpha$ coefficients controlling species' intercepts were drawn from 
Gaussian distributions with unit variance and a mean given by `mean_alpha`.
In the R code at the end of this document, `mean_alpha` is defined as $-1$ 
for presence-absence simulations and $5$ for the abundance simulations.

@@

```{r, cache=TRUE}
make_coefficients = function(n_spp, p_neg, mean_alpha){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2), rate = 1)
